Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-01-04 E-Hypergraphs-LICS24.tex
Section 3.2, moved theorems to the appendix

A typical DPO square looks as follows,  where the two marked squares are pushouts. 
\[
\begin{tikzcd}
L \arrow[d, "m"'] & K \arrow[d] \arrow[l, "f"'] \arrow[r, "g"] & R \arrow[d]   \\
G^{\urcorner}     & C \arrow[l] \arrow[r]                      & ^{\ulcorner}H
\end{tikzcd}
\]
Morphisms $K \to C \to G$ are called a \textit{pushout complement} and morphism $m$ is called a \textit{match},  defined when the corresponding pushout complement exists. When $f,g$ are expected to be monomorphisms.   Constructing $C$ can be intuitively understood as finding an occurrence of $L$ in $G$ and removing from this occurrence everything that has no pre-image in $K$ and then inserting in there everything from $R$ that has no pre-image in $K$ -- hence the two pushouts: one for deletion and one for insertion.  In this case,  we say that $G$ is rewritten into $H$ via the rewrite rule $L \xleftarrow{} K \xrightarrow{} R$.

However, this notion of DPO does not account for hypergraphs with interfaces, i.e. those from $\MdaCospans$ where $l,r$ from $\langle l, r \rangle$ are cospans. To support this we first assume that we only need to maintain the interfaces of $i \xrightarrow{} L \xleftarrow{} j$ and $i \xrightarrow{} R \xleftarrow{} j$ during rewriting, then we can take $K \cong i + j$, i.e. take $K$ to be the coproduct of $i,j$ (remark 3.14~\cite{Frobenius}). What remains is to account for the interfaces in $G$, $C$, and $H$, which brings us to the following modified DPOI square. $J \cong n + m$ where $n$ and $m$ are the input and output interfaces of $G$, $C$, and $H$ respectively. The only difference from DPO square above is that we now explicitly require that the initial object $G$, intermediate object $C$ and final object $H$ all share the same interfaces.

CHRIS: DEFINE DPOI FORMALLY,  IN DEFINITION ENVIRONMENT? PERHAPS COMBINE WITH BOUNDARY COMPLEMENT BELOW,  TWO FIGURES ON ONE LINE? 
\[
\begin{tikzcd}
L \arrow[d, "m"'] & K \arrow[d] \arrow[l, "f"'] \arrow[r, "g"] & R \arrow[d]   \\
G^{\urcorner}     & C \arrow[l] \arrow[r]                      & ^{\ulcorner}H \\
                  & J \arrow[lu] \arrow[u] \arrow[ru]          &              
\end{tikzcd}
\]

To mimic rewrites in $\textsf{PROP}(\Sigma)$ we need to put some restrictions on pushout complements as not all of them lead to mda-cospans. CHRIS: "LEAD TO"? BE PRECISE! :)

\begin{definition}[Boundary complement for $\catname{MACsp_{D}(Hyp_{\Sigma}))}$]
\label{def:boundary_original}
For monogamous cospans $i \xrightarrow{a_1} L \xleftarrow{a_2} j$ and $n \xrightarrow{b_1} G \xleftarrow{b_2} m$
and mono $f : L \to G$, a pushout complement as depicted below

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.4\linewidth]{figures/combinatorial_semantics/boundary_complement.png}
\end{figure}

@@

@@

\begin{definition}[DPOI Rewriting]
Given $G \xleftarrow{} n+m$ and $H \xleftarrow{} n + m$ in $\catname{Hyp_{\Sigma}}$, $G$ rewrites convexly into $H$ with interface $n + m$ --- notation $(G \xleftarrow{} n + m ) \Rrightarrow_{\mathcal{R}}  (H \xleftarrow{} n + m )$ in $\catname{HypI(\Sigma)}$ --- if there exist rule $L \xleftarrow{} i + j \xrightarrow{} R$ in $\mathcal{R}$ and object $C$ and cospan arrows $i+j \xrightarrow{} C \xleftarrow{} n+m$ in $\catname{Hyp_{\Sigma}}$ such that the DPOI diagram above commutes and its marked squares are pushouts and the following conditions hold

@@

@@

@@

\begin{proposition}
    The boundary complement in~\ref{def:boundary_new} when exists is unique
\end{proposition}

@@

>>>>

A typical DPO square looks as follows,  where the two marked squares are pushouts. 
\[
\begin{tikzcd}
L \arrow[d, "m"'] & K \arrow[d] \arrow[l, "f"'] \arrow[r, "g"] & R \arrow[d]   \\
G^{\urcorner}     & C \arrow[l] \arrow[r]                      & ^{\ulcorner}H
\end{tikzcd}
\]
Morphisms $K \to C \to G$ are called a \textit{pushout complement} and morphism $m$ is called a \textit{match},  defined when the corresponding pushout complement exists. When $f,g$ are expected to be monomorphisms, constructing $C$ can be intuitively understood as finding an occurrence of $L$ in $G$ and removing from this occurrence everything that has no pre-image in $K$ and then inserting in there everything from $R$ that has no pre-image in $K$ -- hence the two pushouts: one for deletion and one for insertion.  In this case,  we say that $G$ is rewritten into $H$ via the rewrite rule $L \xleftarrow{} K \xrightarrow{} R$.

To express the rewriting of (hyper-)graphs with interfaces, e.g. the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout with \textit{interfaces} rewriting (DPOI).

\begin{definition}[DPOI rewriting]
Fix a category $\mathbb{C}$ with pushouts.
Given morphisms $G \xleftarrow{} J$ and $H \xleftarrow{} J$ in $\mathbb{C}$, $G$ rewrites into $H$ with interface $J$ if there exists a rewrite rule $L \xleftarrow{} K \xrightarrow{} R$ and object $C$ with suitable morphisms in $\mathbb{C}$ such that the diagram below commutes and two marked squares are pushouts.
\[
\begin{tikzcd}
L \arrow[d, "m"'] & K \arrow[d] \arrow[l, "f"'] \arrow[r, "g"] & R \arrow[d]   \\
G^{\urcorner}     & C \arrow[l] \arrow[r]                      & ^{\ulcorner}H \\
                  & J \arrow[lu] \arrow[u] \arrow[ru]          &              
\end{tikzcd}
\]

\end{definition}
When $\mathbb{C} = \catname{Hyp_{\Sigma}}$, the rewrite rules become cospans of hypergraphs with matching interfaces: $i \xrightarrow{} L \xleftarrow{} j$, $i \xrightarrow{} R \xleftarrow{} j$.
By noting that a cospan $0 \xrightarrow{} L \xleftarrow{[f_{i},f_{j}]} i + j$ encodes the same data as a cospan $i \xrightarrow{f_{i}} L \xleftarrow{f_{j}} j$ (Remark 3.14~\cite{Frobenius}) we can formulate a rewrite rule as a single span $L \xleftarrow{} K \xrightarrow{} R$ where $K \cong i + j$.
Likewise, for the target cospan $n \xrightarrow{} G \xleftarrow{} m$  $J \cong n + m$.
% However, this notion of DPO does not account for hypergraphs with interfaces, i.e. those from $\MdaCospans$ where $l,r$ from $\langle l, r \rangle$ are cospans. To support this we first assume that we only need to maintain the interfaces of $i \xrightarrow{} L \xleftarrow{} j$ and $i \xrightarrow{} R \xleftarrow{} j$ during rewriting, then we can take $K \cong i + j$, i.e. take $K$ to be the coproduct of $i,j$ (remark 3.14~\cite{Frobenius}). What remains is to account for the interfaces in $G$, $C$, and $H$, which brings us to the following modified DPOI square. $J \cong n + m$ where $n$ and $m$ are the input and output interfaces of $G$, $C$, and $H$ respectively. The only difference from DPO square above is that we now explicitly require that the initial object $G$, intermediate object $C$ and final object $H$ all share the same interfaces.

CHRIS: DEFINE DPOI FORMALLY,  IN DEFINITION ENVIRONMENT? PERHAPS COMBINE WITH BOUNDARY COMPLEMENT BELOW,  TWO FIGURES ON ONE LINE?

To mimic rewrites in $\textsf{PROP}(\Sigma)$ with the help of DPOI rewriting in $\catname{Hyp_{\Sigma}}$, we need to put some restrictions both on pushout complement and on matching as not all of them correspond to valid rewrites in $\textsf{PROP}(\Sigma)$. CHRIS: "LEAD TO"? BE PRECISE! :)

\begin{definition}[Boundary complement]
\label{def:boundary_original}
For monogamous cospans $i \xrightarrow{a_1} L \xleftarrow{a_2} j$ and $n \xrightarrow{b_1} G \xleftarrow{b_2} m$
and mono $f : L \to G$, a pushout complement as depicted below

% \begin{figure}[h!]
%     \centering
%     \includegraphics[width=0.4\linewidth]{figures/combinatorial_semantics/boundary_complement.png}
% \end{figure}

% https://q.uiver.app/#q=WzAsNSxbMCwwLCJMIl0sWzIsMCwiaStqIl0sWzAsMiwiR157XFx1cmNvcm5lcn0iXSxbMiw0LCJuK20iXSxbMiwyLCJMXntcXGJvdH0iXSxbMyw0LCJbZF8xLGRfMl0iLDJdLFszLDIsIltiXzEsYl8yXSJdLFswLDIsImYiLDJdLFsxLDAsImEgPSBbYV8xLGFfMl0iLDJdLFsxLDQsImMgPSBbY18xLGNfMl0iXSxbNCwyLCJnIl1d
\[\begin{tikzcd}
	L && {i+j} \\
	\\
	{G^{\urcorner}} && {L^{\bot}} \\
	\\
	&& {n+m}
	\arrow["{[d_1,d_2]}"', from=5-3, to=3-3]
	\arrow["{[b_1,b_2]}", from=5-3, to=3-1]
	\arrow["f"', from=1-1, to=3-1]
	\arrow["{a = [a_1,a_2]}"', from=1-3, to=1-1]
	\arrow["{c = [c_1,c_2]}", from=1-3, to=3-3]
	\arrow["g", from=3-3, to=3-1]
\end{tikzcd}\]

@@

\begin{remark}
    Note how the requirement for $i \xrightarrow{} L \xleftarrow{} j$ and $n \xrightarrow{} G \xleftarrow{} m$ being monogamous, i.e. morphisms in $\MdaCospans{\Sigma}$, is a part of the above definition.
\end{remark}

@@

\begin{definition}
Given $G \xleftarrow{} n+m$ and $H \xleftarrow{} n + m$ in $\catname{Hyp_{\Sigma}}$, $G$ rewrites convexly into $H$ with interface $n + m$ --- notation $(G \xleftarrow{} n + m ) \Rrightarrow_{\mathcal{R}}  (H \xleftarrow{} n + m )$ --- if there exist rule $L \xleftarrow{} i + j \xrightarrow{} R$ in $\mathcal{R}$ and object $C$ and cospan arrows $i+j \xrightarrow{} C \xleftarrow{} n+m$ such that the DPOI diagram above commutes and its marked squares are pushouts and the following conditions hold

@@

ALEKSEI: THE WORD `CONVEXLY' ABOVE MEANS THAT THE REWRITING IS FORMULATED FOR MOMOGAMOUS MORPHISMS. SHALL WE STICK TO THIS FORMULATION?

@@

\begin{theorem}[Theorem~\ref{th:existence_of_pushouts} restatement]
Under certain conditions pushouts in $\catname{EHyp_{\Sigma}}$ exist.   
\end{theorem}

@@

\begin{proposition}[Proposition~\ref{prop:boundary_complement} restatement]
    The boundary complement in~\ref{def:boundary_new} when exists is unique.
\end{proposition}

@@

\begin{theorem}[Existence of pushouts in $\catname{Hyp_{\Sigma}}$]

    Pushouts in $\catname{Hyp_{\Sigma}}$ always exist.
\end{theorem}

\begin{proof}
    Consider a diagram below
    % https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAC0QBfU9TXfIRQBGclVqMWbABrdeIDNjwEiZYePrNWiEAE05fJYKKj11TVJ0AdK3ABmAJzoBjYNIDUursBvYAtgAEAEpcBgr8ykLIAEyk0RqS2iAAitziMFAA5vBEoI4QfkhkIDgQSKISWmx2YfmFiMWlSLGVliCZINQMdABGMAwAChHGOg5YmQAWOLUOBeXUTYgAzOaJbL6OLsBYAPrC3r5YgSEzc4gtiyutSRtOrrvRB7ZHwaE8ebP1F2WIACzUfTAUCQAFolsULEkAFY7aKdEDdPqDYYqUbjKaneoVRb-ECA4HLCFrHQw4SYpBXHGrKo6JhpLhAA

\[
        \begin{tikzcd}
        Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{i_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
        Y \arrow[r, "\sfrac{i_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                        &                                                                  & Q
        \end{tikzcd}
\]

Pushout of two hypergraphs $X = \{V_{X}, E_{X}, s_{X}, t_{X}\}$ and $Y = \{V_{Y}, E_{Y}, s_{Y}, t_{Y}\}$ (we will omit labelling functions here since they are irrelevant) along $Z$ is computed in two steps.
First, a coproduct of $X$ and $Y$ is taken which is 
\[
    X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X + Y}, t_{X+Y}\}
\]
where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $\langle s'_{X}, s'_{Y} \rangle : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};i_{1,V}^{*}$ and $s'_{Y} = s_{Y};i_{2,V}^{*}$ where $i_1,i_2$ are corresponding coproduct injections, similarly for $t_{X+Y}$.
Recall that each arrow in $\catname{Hyp_{\Sigma}}$ is a pair of arrows for edges and nodes and we will further use subscripts $V$ and $E$ when referring to these functions.
Then, consider relations 
\[
    S_{V} = \{(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \; \exists z \in V_{Z} \; . \; x_i = f_{V};i_{1,V}(z) \text{ and } y_j = g_{V};i_{2,V}(z) \text{where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }\}
\]
\[
    S_{E} = \{(x_i,y_j) \in (E_{X} + E_{Y}) \times (E_{X} + E_{Y})\; | \; \exists z \in E_{Z} \; . \; x_i = f_{E};i_{1,E}(z) \text{ and } y_j = g_{E};i_{2,E}(z) \text{where $x_i \in E_{X}$ and $y_j \in E_{Y}$ }\}
\]
and let relations $R_{V}$ and $R_{E}$ be their reflexive symmetric and transitive closures respectively.

\begin{lemma}
    $x R y$ if and only if either $x = y$ or there exists a sequence $(a_1,\ldots,a_{n-1},{a_n})$ such that $x = a_1$ and $y = a_{n}$ such that $a_k S a_{k+1}$ or $a_{k + 1} S a_{k}$ for $k < n$.
\end{lemma}

A quotient of $X+Y$ by these relation is 
\[
    \sfrac{X+Y}{\sim (R_{V},R_{E})} = \{\sfrac{V_{X} + V_{Y}}{\sim R_{V}}, \sfrac{E_{X} + E_{Y}}{\sim R_{E}}, \sfrac{s_{X+Y}}{\sim (R_{V},R_{E})}, \sfrac{t_{X+Y}}{\sim (R_{V},R_{E})}\}
\]
We will then refer to $\sim (R_{V},R_{E})$ just as $\sim$, i.e. writing $\sfrac{V_{X} + V_{Y}}{\sim}$ and the exact relation will be clear from the context.
We have 
\[
    \sfrac{s_{X+Y}}{\sim} : \sfrac{E_{X} + E_{Y}}{\sim} \to (\sfrac{V_{X} + V_{Y}}{\sim})^{*}
\]
In particular, there is an obvious surjective function $[-]_{V} : (V_{X} + V_{Y}) \to (\sfrac{V_{X} + V_{Y}}{\sim})$ that maps elements to their equivalence classes and $[-]_{V}^{*}$ is its extension to sequences. 
Let $s_2 : E_{X} + E_{Y} \to (\sfrac{V_{X} + V_{Y}}{\sim})^{*} = s_{X+Y};[-]^{*}$ and then $\sfrac{s_{X+Y}}{\sim}([e]) = s_{X+Y};[-]^{*}(e) = [s_{X+Y}(e)]^{*}$.
There is also $[-] : E_{X} + E_{Y} \to \sfrac{E_{X} + E_{Y}}{\sim}$ and we will omit subscripts as the correct type will be clear from the argument.
We will also use subscripts when it is important to tell if an element of $E_{X} + E_{Y}$ has in image in either $E_{X}$ or $E_{Y}$ by writing $e_{x}$ or $e_{y}$.
We need to check that if $t_1 \sim t_2$ then $\sfrac{s_{X+Y}}{\sim}([t_1]) = \sfrac{s_{X+Y}}{\sim}([t_2])$. 
There are three options
\begin{enumerate}
    \item $t_1 = e_{y}^{1}$ and $t_2 = e_{y}^{2}$
    \item $t_1 = e_{x}^{1}$ and $t_2 = e_{x}^{2}$
    \item $t_1 = e_{x}$ and $t_2 = e_{y}$ 
\end{enumerate}
We will prove the last point (all the others are analogous).

According to the lemma above $e_{x} \sim e_{y}$ gives us two options.
\begin{enumerate}
    \item Either $e_{x} = e_{y}$
    \item or there is a sequence $w = (a_1, \ldots, a_{n})$ such that $a_1 = e_{x}$ and $a_{n} = e_{y}$.
\end{enumerate}
In the first case the equality holds on the nose.
We then prove the second case by induction on the length of $w$.
\begin{itemize}
    \item In the case $|w| = 2$, $e_{x} S e_{y}$ (or $e_{y} S e_{x}$) which means that there exists $z$ in $E_{Z}$ such that $f_{E};i_{1,E}(z) = e_{x}$ and $g_{E};i_{2,E}(z) = e_{y}$.
            Recall that because $f$ and $g$ are homomorphisms, the following equalities hold
            \[
            s_{Z};f_{V}^{*};i_{1,V}^{*}(z) = f_{E};i_{1,E};s_{X+Y}(z) = s_{X+Y}(e_{x})
            \]
            \[
            s_{Z};g_{V}^{*};i_{2,V}^{*}(z) = g_{E};i_{2,E};s_{X+Y}(z) = s_{X+Y}(e_{y})
            \]
            Then,
            \[
            \sfrac{s_{X+Y}}{\sim}([e_{x}]) = [s_{X+Y}(e_{x})]^{*} = [s_{Z};f_{V}^{*};i_{1,V}^{*}(z)]^{*}
            \]
            and
            \[
            \sfrac{s_{X+Y}}{\sim}([e_{y}]) = [s_{X+Y}(e_{y})]^{*} = [s_{Z};g_{V}^{*};i_{2,V}^{*}(z)]^{*}
            \]
            Then, by noting that $f_{V};i_{1,V}(z) = g_{V};i_{2,V}(z)$ for all $z$ in $V_{Z}$ (because $\sfrac{V_{X} + V_{Y}}{\sim}$ is so defined) we have $[s_{X+Y};(e_{x})]^{*} = [s_{X+Y}(e_{y})]^{*}$.
    \item Suppose if $e_{x} S e_{y}$ by a chain of length $n$ or less then $[s_{X+Y}(e_{x})]^{*} = [s_{X+Y}(e_{y})]^{*}$.
    \item Now suppose $e_{x} S e_{y}$ by a chain $w = (a_1, \ldots, a_{n+1})$ and a subchain $(a_1, \ldots, a_{n})$ satisfies the inductive hypothesis and hence $s_{X+Y};s_{1}^{*}(e_{x}) = s_{X+Y};s_{1}^{*}(a_{n})$.
            Note that $S_{E}, S_{V}$ relate only elements with the pre-image in $E_{X}$ (respectively, $V_{X}$) with elements with the pre-image in $E_{Y}$ (respectively, $V_{Y}$). 
            Elements with the pre-image in the same set are not related.
            Then we have $a_n S a_{n+1} = e_{y}$ and need to show that $[s_{X+Y}(a_{n+1})]^{*} = [s_{X+Y}(e_{y})]^{*}$.
            This follows by exactly the same argument as the base case as $a_{n+1}$ is necessarily in $E_{X}$.
            And we finally have $[s_{X+Y}(e_{x})]^{*} = [s_{X+Y}(a_{n+1})]^{*} = [s_{X+Y}(e_{y})]^{*}$.
\end{itemize}

Next 
\[
    \sfrac{i_{1,V}}{\sim} : V_{X} \to \sfrac{V_{X} + V_{Y}}{\sim} = [i_{1,V}]
\]
and
\[
\sfrac{i_{1,E}}{\sim} : E_{X} \to \sfrac{E_{X} + E_{Y}}{\sim} = [i_{1,E}]
\]
similarly for $\sfrac{i_{2,E}}{\sim}$ and $\sfrac{i_{2,V}}{\sim}$
We need to show
\begin{enumerate}
    \item $s_{X};(\sfrac{i_{1,V}}{\sim})^{*}(e) = \sfrac{i_{1,E}}{\sim};\sfrac{s_{X+Y}}{\sim}(e)$
    \item $t_{X};(\sfrac{i_{1,V}}{\sim})^{*}(e) = \sfrac{i_{1,E}}{\sim};\sfrac{t_{X+Y}}{\sim}(e)$
\end{enumerate}
and likewise for $i_{2}$.
Let's check the first one for which we first need to check that $([-]_{E},[-]_{V})$ is a homomorphisms, i.e.
\[
[s_{X+Y}(e)]^{*} = \sfrac{s_{X+Y}}{\sim}([e]) = [s_{X+Y}(e)]^{*}
\]
hence, $(\sfrac{i_{1,V}}{\sim}, \sfrac{i_{1,E}}{\sim})$ is a homomorphism because $(i_{1,V},i_{1,E})$ and $([-]_{E},[-]_{V})$ are.

Now, the square obviously commutes.
To show that $\sfrac{X + Y}{\sim}$ is a pushout we need to argue that for each $Q$ and morphisms $j_1 = (j_{1,V}, j_{1,E})$ and $j_2 = (j_{2,V}, j_{2,E})$ such that the above square commutes there is a unique $u = (u_{V}, u_{E})$ such that $j_1 = \sfrac{i_{1}}{\sim} ; u$ and $j_2 = \sfrac{i_2}{\sim} ; u$.
We will define such $u$ as $u_{E}([i_{1,E}(e_{x})]) = j_{1,E}(e_{x})$ and $u_{E}([i_{2,E}(e_{y})]) = j_{2,E}(e_y)$ (similarly for $u_{V}$) and claim that it is a well-defined function by showing that if $t_1 \sim t_2$ then $u_{E}(t_1) = u_{E}(t_2)$.
There are three options analogoues to the proof above, so we will show the base case for $t_1 = i_{1,E}(e_{x})$ and $t_2 = i_{2,E}(e_{y})$.
$t_1 \sim t_2$ implies that there exists at least one $z$ in $E_{Z}$ such that $i_{1,E}(e_{x}) = f_{E};i_{1,E}(z)$ (or $e_{x} = f_{E}(z)$) and $[i_{1,E}(e_{x})] = \sfrac{i_{1,V}}{\sim}(e_x)$, and $e_y = g_{E}(z)$ $[in_{2,E}(e_{y})] = \sfrac{i_{2,V}}{\sim}(e_y)$.
By applying $j_{1,E}$ to the first we get $j_{1,E}(f_{E}(z)) = j_{1,E}(e_x)$ and similarly for $j_{2,E}$: $j_{2,E}(g_{E}(z)) = j_{2,E}(e_y)$ and $j_{1,E}(e_{x}) = j_{2,E}(e_{y})$ by commutativity.
Hence, $u_{E}([in_{1,E}(e_x)]) = u_{E}([in_{2,E}(e_y)])$.

Next, $u$ is a homomorphism because $j_1$ and $j_2$ are.
Finally, $u$ is unique.
Assume, that there exists $w : \sfrac{X+Y}{\sim R} \to Q$ for the same $j_1$ and $j_2$.
In order for them to commute $w_{E}([i_{1,E}(x)]) = j_1(x)$ and $w([in_{2,E}(y)]) = j_2(y)$ (similarly for $w_{V}$) which is the same as $u_{E}$.
That is, $j_1$ and $j_2$ determine $u$ uniquely.
\end{proof}

\begin{remark}
    Both relations can be considered as functions defined on $V_{\mathcal{F}} + E_{\mathcal{F}}$, i.e. on the coproduct of nodes and edges.
    To make things well-typed, we will use corresponding coproduct injections $i_{V_{\mathcal{F}}} \to V_{\mathcal{F}} + E_{\mathcal{F}}$ and $i_{E_{\mathcal{F}}} \to V_{\mathcal{F}} + E_{\mathcal{F}}$ when passing either node or edge into these functions.
    For example, an immediate successor of a node $x$ can be written functionally as $<_{c,\mathcal{F}}^{\mu}(i_{V_{\mathcal{F}}}(x))$.
\end{remark}

\begin{remark}
    Using functional notation we can reformulate the notion of homomorphism of two e-hypergraphs in the following way.
    \begin{definition}{E-hypergraphs homomorphism}
        \label{def:e-homo-2}    
        A homomorphism $\phi: \mathcal{F} \to \mathcal{G}$ of e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that

        \begin{enumerate}
            \item $\phi$ is hypergraph homomorphism

            \item \[
                \phi_{E}(<_{c,\mathcal{F}}^{\mu}(i_{V_{\mathcal{F}}}(x))) = <_{c,\mathcal{G}}^{\mu}(\phi_{V};i_{V_{\mathcal{G}}}(x))
                \] when $x$ is a node,
                and
                \[
                \phi_{E}(<_{c,\mathcal{F}}^{\mu}(i_{E_{\mathcal{F}}}(x))) = <_{c,\mathcal{G}}^{\mu}(\phi_{E};i_{E_{\mathcal{G}}}(x))  
                \] when $x$ is an edge.
                We do not require such an equation when $[x) = \varnothing$.

            % \item
            % $x_1 \#_{\mathcal{F}} x_2$ iff $\phi(x_1) \#_{\mathcal{G}} \phi(x_2)$, where $x_1, x_2 \in \mathcal{F}$

                \item
            When $x \in E_{\mathcal{F}}$
            \[
                \langle \phi_{V};i_{V_{\mathcal{G}}}, \phi_{E};i_{E_{\mathcal{G}}} \rangle^{*}(\consistency_{\mathcal{F}}^{\mu}(i_{E_{\mathcal{F}}}(x)))
                \subseteq
                \consistency_{\mathcal{G}}^{\mu}(\phi_{E};i_{E_{\mathcal{G}}}(x))
            \]
            , where $\phi_{V};i_{V_{\mathcal{G}}} : V_{\mathcal{F}} \to V_{\mathcal{G}} + E_{\mathcal{G}}$, and similarly for $\phi_{E};i_{E_\mathcal{G}}$ so that $\langle \phi_{V};i_{V_{\mathcal{G}}}, \phi_{E};i_{E_{\mathcal{G}}} \rangle : V_{\mathcal{F}} + E_{\mathcal{F}} \to  V_{\mathcal{G}} + E_{\mathcal{G}}$.
            And, when $x \in V_{\mathcal{F}}$
            \[
                \langle \phi_{V};i_{V_{\mathcal{G}}}, \phi_{E};i_{E_{\mathcal{G}}} \rangle^{*}(\consistency_{\mathcal{F}}^{\mu}(i_{V_{\mathcal{F}}}(x)))
                \subseteq
                \consistency_{\mathcal{G}}^{\mu}(\phi_{V};i_{V_{\mathcal{G}}}(x))
            \]
            \end{enumerate}
    \end{definition}
\end{remark}

\begin{theorem}[Existence of pushouts in $\catname{EHyp_{\Sigma}}$]
\label{th:existence_of_pushouts}
Consider the following span in $\catname{EHyp_{\Sigma}}$
% https://q.uiver.app/#q=WzAsMyxbMCwwLCJaIl0sWzIsMCwiWCJdLFswLDIsIlkiXSxbMCwxLCJmIl0sWzAsMiwiZyIsMl1d
\[\begin{tikzcd}
	Z && X \\
	\\
	Y
	\arrow["f", from=1-1, to=1-3]
	\arrow["g"', from=1-1, to=3-1]
\end{tikzcd}\]
such that
\begin{enumerate}
    \item $Z$ is a \textit{descrete} e-hypergraph.
    \item $f,g$ are monos.
    \item $[f_{V}(v_i)) = [f_{V}(v_j))$ and $[g_{V}(v_i)) = [g_{V}(v_j))$ for all $v_{i},v_{j}$ in $V_{Z}$.
    \item If $[f_{V}(v)) \not = \varnothing$ then $[g_{V}(v)) = \varnothing$ and if $[g_{V}(v)) \not = \varnothing$ then $[f_{V}(v)) = \varnothing$.
    \item $\consistency^{\mu}(f(v_i)) = \consistency^{\mu}(f(v_j))$ and $\consistency^{\mu}(g(v_i)) = \consistency^{\mu}(g(v_j))$ for all $v_i,v_j$ in $V_{Z}$.
\end{enumerate}    
then the pushout $X +_{f,g} Y$ exists.
\end{theorem}
\begin{proof}
    Recall the diagram below.
    \[
        \begin{tikzcd}
        Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{i_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
        Y \arrow[r, "\sfrac{i_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                        &                                                                  & Q
        \end{tikzcd}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, <_{c,X+Y} \}
    \]

    once again omitting the labelling function and where source and target functions are as before.
    And consider the same relations $S_{V},S_{E}$ and $R_{V}, R_{E}$.

    We then quotient the sets of edges and nodes in $X + Y$ by $R_{V},R_{E}$
    \[
        \sfrac{X + Y}{\sim (R_{V},R_{E})} = \{\sfrac{V_{X} + V_{Y}}{\sim R_{V}}, \sfrac{E_{X} + E_{Y}}{\sim R_{E}}, \sfrac{s_{X+Y}}{\sim (R_{V},R_{E})}, \sfrac{t_{X,Y}}{\sim (R_{V},R_{E})}, \#_{\sfrac{X+Y}{\sim (R_{V},R_{E})}}, <_{c,\sfrac{X + Y}{\sim (R_{V},R_{E})}} \}    
    \]
    where quotienting on source and target functions is taken as in $Hyp_{\Sigma}$ and we will further omit writing the relations explicitly.
    Recall that $<_{c}$ is essentially a transitive clousure of $<_{c}^{\mu}$ and hence we can first define 
    \[<_{c,\sfrac{X+Y}{\sim}}^{\mu}(i_{\sfrac{E_{X} + E_{Y}}{\sim}}[u]_{E})
    \] and 
    \[
        <_{c,\sfrac{X+Y}{\sim}}^{\mu}(i_{\sfrac{V_{X} + V_{Y}}{\sim}}[u]_{V})
    \] where $i_*$ are injections into $\sfrac{V_{X} + V_{Y}}{\sim} + \sfrac{E_{X} + E_{Y}}{\sim}$.
    We will consider several cases.
    Assume that $u$ is an element of $V_{X+Y}$ and
    \begin{enumerate}
        % \item $u$ has a pre-image $z$ in $V_{Z}$, $[z) \not = \varnothing$.
        %       The last requirement means that there necessarily exist $x = <_{c,X}^{\mu}(f_{E}(z)) = f_{E}(<_{c,Z}^{\mu}(z))$ and $y = <_{c,Y}^{\mu}(g_{E}(z)) = g_{E}(<_{c,Z}^{\mu}(z))$.
        %       We then define $<_{c,\sfrac{X+Y}{\sim}}^{\mu}([u]) = [<_{c,X+Y}^{\mu}(u)]$
        \item $u$ has a pre-image $z$ in $V_{Z}$ (by discreteness, $[z) = \varnothing$) and $[f_{V}(z)) = \varnothing$ $[g_{V}(z)) \not = \varnothing$.
              The last condition means that there exists $e_y = <_{c,Y}^{\mu}(i_{V_{Y}}(g_{V}(z)))$.
              We then define 
              \[
                <_{c,\sfrac{X+Y}{\sim}}^{\mu}(i_{\sfrac{V_{X} + V_{Y}}{\sim}}[u]) = [g_{V};i_{V_{Y}};<_{c,Y}^{\mu};i_{2,E}(z)] = [i_{2,E}(e_y)]
              \]
        \item $u$ has a pre-image $z$ in $V_{Z}$, (by discreteness, $[z) = \varnothing$) and $[f_{V}(z)) \not = \varnothing$ $[g_{V}(z)) = \varnothing$.
              The one before the last condition means that there exists $e_x = <_{c,X}^{\mu}(i_{V_{X}}(f_{V}(z)))$.
              We then define 
              \[
                <_{c,\sfrac{X+Y}{\sim}}^{\mu}(i_{\sfrac{V_{X} + V_{Y}}{\sim}}[u]) = [f_{V};i_{V_{X}};<_{c,X}^{\mu};i_{1,E}(z)] = [i_{1,E}(e_x)]
              \]
        \item $u$ has no pre-image in $Z$ and $<_{c,X+Y}^{\mu}(i_{V_{X} + V_{Y}}(u)) = e'$
        \[
            <_{c,\sfrac{X+Y}{\sim}}^{\mu}(i_{\sfrac{V_{X} + V_{Y}}{\sim}}[u]) = [<_{c,X+Y}^{\mu}(i_{V_{X} + V_{Y}}(u))] = [e']
        \]
        \item $u$ has no pre-image in $Z$ and $<_{c,X+Y}^{\mu}(i_{V_{X} + V_{Y}}(u))$ is undefined (i.e. $[u) = \varnothing$), but there exists $[e]$ such that $<_{c,\sfrac{X+Y}{\sim}}^{\mu}(i_{\sfrac{E_{X} + E_{Y}}{\sim}}([e])) = [e']$ and $[u] \in t([e])$ or $[u] \in s([e])$, then
              \[ 
                <_{c,\sfrac{X+Y}{\sim}}^{\mu}(i_{\sfrac{V_{X} + V_{Y}}{\sim}}[u]) = <_{c,\sfrac{X+Y}{\sim}}^{\mu}(i_{\sfrac{E_{X} + E_{Y}}{\sim}}[e])
              \]
              Assumptions above guarantee that the predecessor of such node does not depend on the choice of $e$ if there are several of them.
    \end{enumerate}
    % \question{In the case when images of $f$ and $g$ are top-level the function is undefined. Shall we introduce bottom?}
    Now, if $u$ is an element of $E_{X+Y}$.
    We have the only case of $u$ having no pre-image in $E_{Z}$ because it is empty.
              Consider two cases.
              \begin{enumerate}
